### 2021-04-23 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
Made some changes to address points in comments pdf

On the unit interval, $[-1,1]$, we note that there is a hierarchy of orthogonal polynomials (OPs) in the sense that \cite[table 18.3.1, eqn 18.9.15]{DLMF}:
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x) \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x)
}
where $P^{(a,b)}_l (x)$ is the $l$ degree \textit{Jacobi polynomial}, and $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)} \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'}.
}
Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Jacobi polynomial: \bstodo{Sheehan - do you know what citation I can use for the definition of the Associated Legendre Polynomials (as the mth derivative of the ordinary Legendre Polynomials)?}
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x)
}
for $m = -l, \dots, -1$. Here, $n! := n \: (n-1) \: (n-2) \dots 1$ for $n \in \N_0$ is the standard factorial, with $0! = 1$. Although referred to as the associated Legendre \enquote{polynomials}, they are not strictly polynomials when $m$ is odd, and as such some authors refer to them as associated Legendre functions instead. Moreover, it should be noted too that the inclusion of the $(-1)^m$ factor in the definition for $m > 0$, known as the \textit{Condon-Shortley phase} in physics, is sometimes omitted.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this document we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi.
}
where $\rho(z) := \sqrt{1-z^2}$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}. \label{eqn:clmdef}
} 
The spherical harmonics as defined are then orthonormal with respect to the inner product
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$.

@@

\begin{proof}
	The expression for multiplication of the spherical harmonic $\Ylm$ by $z$ is simply a consequence of the third result in \bsrefcor{corollary:jacobirelations}:
\bseqn{
	z\,\Ylmxyz &= \clm \eimtheta \rho(z)^{|m|} z P^{(|m|,|m|)}_{l-|m|}(z)  \\
		     &= \clm \eimtheta \rho(z)^{|m|} \big[ \tilde \gamma_{l,m,1} P^{(|m|,|m|)}_{l-|m|-1}(z) + \tilde \gamma_{l,m,2} P^{(|m|,|m|)}_{l-|m|+1}(z) \big] \\
		     &= \gamma_{l,m,1} Y^{m}_{l-1}(x,y,z) + \gamma_{l,m,2} Y^{m}_{l+1}(x,y,z).
}

@@

These recurrences lead to Jacobi operators that correspond to multiplication by $x,y,z$. Define for $l \in \No$:
\bseqn{
	\SHvecl := 
		\begin{pmatrix}
			Y^{-l}_l \\
			\vdots \\
			Y^l_l
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\SHvec := 
		\begin{pmatrix}
			\SHvec_0 \\
			\hline
			\SHvec_1 \\
			\hline
			\SHvec_2 \\
			\vdots \\
		\end{pmatrix}.
}
and let the Jacobi matrices $J^x, J^y, J^z$ be given by 
\bseqnnumber{
	J^x \SHvec = x \SHvec, \quad J^y \SHvec = y \SHvec, \quad J^z \SHvec = z \SHvec \label{eqn:jacobimatsdef}.
}
The Jacobi matrices act on the coefficients vector of a function's expansion in the spherical harmonic basis. For example, let the function $f(x,y,z) : \Omega \to \C$ be approximated by its expansion $f(x,y,z) = \SHvec(x,y,z)^\top \: \fvec$ up to some degree order $N$. Then $x f(x,y,z)$ is approximated by $\SHvec(x,y,z)^\top \: J^x \: \fvec$, i.e. $ J^x \: \fvec$ is the coefficients vector for the expansion of the function $(x,y,z) \mapsto x \: f(x,y,z)$ in the spherical harmonics basis.

@@

%
%\subsection{Laplacian operator}
%
%The spherical harmonics are named as such because they are indeed harmonic functions -- that is, they satisfy $\nabla^2 \Ylm = \lambda \Ylm$ for some eigenvalue $\lambda$. It can be shown that in fact, for any $l \in \No$, $-l \le m \le l$, \bstodo{cite}
%\bseqn{
%	\nabla^2 \Ylm = - l (l+1) \Ylm.
%}
%This allows us to write down a sparse (in fact, diagonal) matrix operator for the Laplacian. Recall that a function $f(x,y,z)$ defined on the unit sphere $\Omega$ is approximated by its expansion $\SHvec^\top \fvec$ for some coefficients vector $\fvec$ up to degree order $N \in \N$. Then the function $(x,y,z) \mapsto \nabla^2 f(x,y,z)$ is approximated by $\SHvec^\top \Delta \fvec$ where $\Delta$ is the matrix operator is then simply given by
%\bseqn{
%	\Delta := 
%		\begin{pmatrix}
%			\Delta_0 & & & \\
%			& \Delta_1 & & \\
%			& & \ddots & \\
%			& & & \Delta_N
%		\end{pmatrix}, \quad
%	\Delta_l := - l (l+1) \idmat{2l+1} \in \R^{(2l+1)\times(2l+1)}.
%}

@@

\bstodoinline{Is this paragraph enough? Or do I need to put more detail into explaining these methods, or derive my own?}
In spectral space, we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \int_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta
}
using the orthonormality of the spherical harmonics. Methods to calculate these coefficients exist, known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}).

@@

We start as before for the scalar case by finding $x \, \gradYlm (x,y,z)$, $x \,\gradpYlm (x,y,z)$ etc. in terms of $\gradYlmprime (x,y,z)$, $\gradpYlmprime (x,y,z)$. A lovely consequence of our definitions of the scalar and vector spherical harmonics is that we can once again find explicit expressions for the coefficients in the aforementioned relations. To this end, we recall and derive some important relations of the complex exponential and Jacobi polynomials that are a part of the spherical harmonics definitions. In particular we first note that:
\bseqnnumber{
	\int_0^{2\pi} \: \eimtheta \: e^{-im'\theta} \: \cos(\theta) \D\theta &= \pi (\delta_{m',m-1} + \delta_{m',m+1}) \label{eqn:intcomplexexpcos} \\
	\int_0^{2\pi} \: \eimtheta \: e^{-im'\theta} \: \sin(\theta) \D\theta &= i \; \pi (\delta_{m',m-1} - \delta_{m',m+1}) \label{eqn:intcomplexexpsin}
}
and, for $l \in \No, m \in \Z$ s.t. $-l \le m \le l$,
\bseqnnumber{
	&\int_{-1}^1 \: \jaclmabs{l}{m}(z) \: \jaclmabs{l'}{m}(z) \: \rho(z)^{2\abs{m}} \: \D z = \delta_{l',l} \: {1 \over 2\pi (\clm)^2} \label{eqn:normjac} \\
	&\jaclmabsdz{l}{m}(z) = d_{l,m} \: \jaclmbracket{l}{\abs{m} + 1} \quad \text{where } 
	d_{l,m} := 
		\begin{cases}
			\half (l + \abs{m} + 1) & l > \abs{m} \\
			0 & l = \abs{m}
		\end{cases}. \label{eqn:diffjac}
}
\bsrefeqncapital{eqn:normjac} and \bsrefeqn{eqn:diffjac} are simply consequences of the definition of the Jacobi polynomials -- see \cite[Table 18.3.1]{DLMF} and \cite[18.9.15]{DLMF}.

@@

\subsubsection{Obtaining coefficients}
\bstodoinline{As for the scalar case, do I need to say more here about the transforms?}
As eluded to, we would like to be able to obtain the coefficients for the expansion of a vector valued function in the VSH basis up to a given degree, similar to how we approach the scalar case. To do this, we can use a vector spherical harmonic transform. The development of vector spherical harmonic transforms is well established \cite{swarztrauber1993vector}, and continues today (see \cite{gia2019favest}).

@@

Let $\uvec(x,y,z)$ be the tangential velocity of a flow and $h(x,y,z)$ be the height deviation of the flow from some constant reference height $\calH$. Define $\rvec$ as the unit outward normal vector at the point on the sphere $(x,y,z)$, so that
\bseqn{
	\rvec = \begin{pmatrix} x \\ y \\ z \end{pmatrix}.
}
The linear SWEs are
\bseqn{
	\begin{cases}
		\pfpx{t}{\uvec} + f \rvec \times \uvec - \gradS h = \mathbf{0} \\
		\pfpx{t}{h} + \calH \gradS \cdot \uvec = 0
	\end{cases}
}
where $f = 2 \rotrate \cos(\varphi) =  2 \rotrate z$ is the Coriolis parameter, and $\rotrate$ is the rotation rate of the Earth (note that $\varphi = {\pi \over 2} - \text{latitude}$ \bstodo{make a footnote}). On a full scale Earth, the rotation rate is $\rotrate = 7.2921 × 10^{-5}$ rad s$^{-1}$ \bstodo{make a footnote}.

A keen eye may notice that -- while we have operators for the multiplication by z, the spherical gradient and divergence -- we require one more operator not yet defined for the unit vector cross product $\rvec \times \uvec$.
\begin{definition}\label{def:unitveccrossproduct}
	Define the operator matrix $\calR$ according to:
\bseqn{
	\rvec(x,y,z) \times \uvec(x,y,z) &= \VSHvecfull(x,y,z)^\top \: \calR \uvec^c.
}
\end{definition}
\begin{lemma}
	The operator $\calR$ defined in \bsrefdef{def:unitveccrossproduct} is sparse with banded-block-banded structure. More specifically, $\calR$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(1,1)$ (i.e. is tridiagonal).
\end{lemma}
\begin{proof}
	Using the relations that $\gradYlm = \gradS \Ylm$ and $\gradpYlm = \rvec \times \gradS \Ylm$ for each $l \in \N$, $m \in \Z$ s.t. $-l \le m \le l$, we have that:
\bseqn{
	\rvec \times \bold{u} &= \rvec \times \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \gradYlm + u^{\gradpY}_{l,m} \gradpYlm ] \\
	&= \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \rvec \times \gradYlm + u^{\gradpY}_{l,m} \rvec \times ( \rvec \times \gradYlm ) ] \\
	&= \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \: \gradpYlm - u^{\gradpY}_{l,m} \: \gradYlm ) ].
}
We can then write down the operator $\calR$ as follows:
\bseqn{
	\calR &=
		\begin{pmatrix}
			R & & & \\
			& R & & \\
			& & & \ddots
		\end{pmatrix} \in \R^{2(N+1)^2 \times 2(N+1)^2},
}
where the blocks $R$ are simply given by
\bseqn{
	R &:= 
		\begin{pmatrix}
			0 & 1 \\
			-1 & 0
		\end{pmatrix} \in \R^{2\times2}.
}
\end{proof}

>>>>

On the unit interval, $[-1,1]$, we note that there is a hierarchy of orthogonal polynomials (OPs) in the sense that \cite[table 18.3.1, eqn 18.9.15]{DLMF}:
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x) \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x)
}
where $P^{(a,b)}_l (x)$ is the $l$ degree \textit{Jacobi polynomial}, and $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)} \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'}.
}
Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[14.21.1]{DLMF}:
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x)
}
for $m = -l, \dots, -1$. Although often referred to as the associated Legendre \enquote{polynomials} as we do here, they are not strictly polynomials when $m$ is odd, and as such some authors refer to them as associated Legendre functions instead. Moreover, it should be noted too that the inclusion of the $(-1)^m$ factor in the definition for $m > 0$, known as the \textit{Condon-Shortley phase} in physics, is sometimes omitted.  \bstodoinline{I have cited DLMF on ALFs equation that defines the ALFs. Now need to explain how that relates to the definition in terms of mth derivative of normal Legendre poly when l,m are integers}

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this document we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi.
}
where $\rho(z) := \sqrt{1-z^2}$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}. \label{eqn:clmdef}
} 
The spherical harmonics as defined are then orthonormal with respect to the inner product
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in $\R^3$.

@@

The results in \bsreflemma{lemma:shxyzrelations} are fairly involved, and can be derived via \textit{Clebsch-Gordan coefficients}.

\begin{proof}[Proof of \bsreflemma{lemma:shxyzrelations}]
	The expression for multiplication of the spherical harmonic $\Ylm$ by $z$ is simply a consequence of the third result in \bsrefcor{corollary:jacobirelations}:
\bseqn{
	z\,\Ylmxyz &= \clm \eimtheta \rho(z)^{|m|} z P^{(|m|,|m|)}_{l-|m|}(z)  \\
		     &= \clm \eimtheta \rho(z)^{|m|} \big[ \tilde \gamma_{l,m,1} P^{(|m|,|m|)}_{l-|m|-1}(z) + \tilde \gamma_{l,m,2} P^{(|m|,|m|)}_{l-|m|+1}(z) \big] \\
		     &= \gamma_{l,m,1} Y^{m}_{l-1}(x,y,z) + \gamma_{l,m,2} Y^{m}_{l+1}(x,y,z).
}

@@

These recurrences lead to Jacobi operators that correspond to multiplication by $x,y,z$ of the OPs. Define for $l \in \No$:
\bseqn{
	\SHvecl := 
		\begin{pmatrix}
			Y^{-l}_l \\
			\vdots \\
			Y^l_l
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\SHvec := 
		\begin{pmatrix}
			\SHvec_0 \\
			\hline
			\SHvec_1 \\
			\hline
			\SHvec_2 \\
			\vdots \\
		\end{pmatrix}.
}
Note that these OP vectors we have defined are grouped and ordered according to the degree $l$, and within each degree ordered by the Fourier mode $m$. Let the Jacobi matrices $J_x, J_y, J_z$ hence be given by 
\bseqnnumber{
	J_x \SHvec(x,y,z) = x \SHvec(x,y,z), \quad J_y \SHvec(x,y,z) = y \SHvec(x,y,z), \quad J_z \SHvec(x,y,z) = z \SHvec(x,y,z) \label{eqn:jacobimatsdef}
}
for $(x,y,z) \in \Omega$. These Jacobi matrices can act on the coefficients vector of a function's expansion in the spherical harmonic basis to implement the operation for multiplication by $x$, $y$ or $z$. For example, given sufficiently large $n \in \N$ let the function $f(x,y,z) : \Omega \to \C$ be given by its expansion
\bseqn{
	f(x,y,z) = \sum_{l=0}^{N} \sum_{m=-l}^{l} f_{l,m} \; \Ylm(x,y,z) = \SHvec(x,y,z)^\top \: \fvec
}
where 
\bseqn{
	\fvec_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			\vdots \\
			f_{l,-l}
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\fvec := 
		\begin{pmatrix}
			\fvec_0 \\
			\vdots \\
			\fvec_N
		\end{pmatrix}.
}
We say that $\fvec$ is the \textit{coefficients vector} of the function $f$. Then $x f(x,y,z)$ is approximated by $\SHvec(x,y,z)^\top \: J_x^\top \: \fvec$. In other words, $J_x^\top \: \fvec$ is the coefficients vector for the expansion of the function $(x,y,z) \mapsto x \: f(x,y,z)$ in the spherical harmonics basis.

@@

@@

In spectral space, we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \int_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta
}
using the orthonormality of the spherical harmonics. As discussed in \bsrefsection{section:introduction}, methods to calculate these coefficients exist known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}). Additionally, packages are available for public use that implement efficient spherical harmonics transforms, for example the FastTransforms package \cite{slevinsky2019fast}.

@@

We start as before for the scalar case by finding $x \, \gradYlm (x,y,z)$, $y \,\gradYlm (x,y,z)$, $z \, \gradYlm (x,y,z)$, and $x \,\gradpYlm (x,y,z)$, $y \, \gradpYlm (x,y,z)$, $z \,\gradpYlm (x,y,z)$ in terms of $\gradYlmprime (x,y,z)$, $\gradpYlmprime (x,y,z)$. A lovely consequence of our definitions of the scalar and vector spherical harmonics is that we can once again find explicit expressions for the coefficients in the aforementioned relations. To this end, we recall and derive some important relations of the complex exponential and Jacobi polynomials that are a part of the spherical harmonics definitions. In particular we first note that:
\bseqnnumber{
	\int_0^{2\pi} \: \eimtheta \: e^{-im'\theta} \: \cos(\theta) \D\theta &= \pi (\delta_{m',m-1} + \delta_{m',m+1}) \label{eqn:intcomplexexpcos} \\
	\int_0^{2\pi} \: \eimtheta \: e^{-im'\theta} \: \sin(\theta) \D\theta &= i \; \pi (\delta_{m',m-1} - \delta_{m',m+1}) \label{eqn:intcomplexexpsin}
}
and, for $l \in \No, m \in \Z$ s.t. $-l \le m \le l$,
\bseqnnumber{
	&\int_{-1}^1 \: \jaclmabs{l}{m}(z) \: \jaclmabs{l'}{m}(z) \: \rho(z)^{2\abs{m}} \: \D z = \delta_{l',l} \: {1 \over 2\pi (\clm)^2} \label{eqn:normjac} \\
	&\jaclmabsdz{l}{m}(z) = d_{l,m} \: \jaclmbracket{l}{\abs{m} + 1} \quad \text{where } 
	d_{l,m} := 
		\begin{cases}
			\half (l + \abs{m} + 1) & l > \abs{m} \\
			0 & l = \abs{m}
		\end{cases}. \label{eqn:diffjac}
}
\bsrefeqncapital{eqn:normjac} and \bsrefeqn{eqn:diffjac} are simply consequences of the definition of the Jacobi polynomials -- see \cite[Table 18.3.1]{DLMF} and \cite[18.9.15]{DLMF}.

@@

\subsubsection{Obtaining coefficients}

As eluded to, we would like to be able to obtain the coefficients for the expansion of a vector valued function in the VSH basis up to a given degree, similar to how we approach the scalar case. To do this, we can use a vector spherical harmonic transform. The development of vector spherical harmonic transforms is well established \cite{swarztrauber1993vector}, and continues today (see \cite{gia2019favest}).

@@

Let $\uvec(x,y,z)$ be the tangential velocity of a flow and $h(x,y,z)$ be the height deviation of the flow from some constant reference height $\calH$. Define $\rvec$ as the unit outward normal vector at the point on the sphere $(x,y,z)$, so that
\bseqn{
	\rvec = \begin{pmatrix} x \\ y \\ z \end{pmatrix}.
}
The linear SWEs are
\bseqn{
	\begin{cases}
		\pfpx{t}{\uvec} + f \rvec \times \uvec - \gradS h = \mathbf{0} \\
		\pfpx{t}{h} + \calH \gradS \cdot \uvec = 0
	\end{cases}
}
where $f = 2 \rotrate \cos(\varphi) =  2 \rotrate z$ is the Coriolis parameter\footnote{Normally, the Coriolis parameter is expressed in terms of the latitude: $\text{latitude} = {\pi \over 2} - \varphi$}, and $\rotrate$ is the rotation rate of the sphere surface (that is, the amount the body rotates per unit time)\footnote{On a full scale Earth, the rotation rate is $\rotrate = 7.2921 × 10^{-5}$ rad s$^{-1}$.}.

A keen eye may notice that -- while we have operators for the multiplication by $z$, the spherical gradient and divergence -- we require one more operator not yet defined for the unit vector cross product $\rvec \times \uvec$.
\begin{definition}\label{def:unitveccrossproduct}
	Define the operator matrix $\calR$ according to:
\bseqn{
	\rvec(x,y,z) \times \uvec(x,y,z) &= \VSHvecfull(x,y,z)^\top \: \calR \uvec^c.
}
\end{definition}
\begin{lemma}
	The operator $\calR$ defined in \bsrefdef{def:unitveccrossproduct} is sparse with banded-block-banded structure. More specifically, $\calR$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(1,1)$ (i.e. is tridiagonal).
\end{lemma}
\begin{proof}
	Using the relations that $\gradYlm = \gradS \Ylm$ and $\gradpYlm = \rvec \times \gradS \Ylm$ for each $l \in \N$, $m \in \Z$ s.t. $-l \le m \le l$, we have that:
\bseqn{
	\rvec \times \bold{u} &= \rvec \times \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \gradYlm + u^{\gradpY}_{l,m} \gradpYlm ] \\
	&= \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \rvec \times \gradYlm + u^{\gradpY}_{l,m} \rvec \times ( \rvec \times \gradYlm ) ] \\
	&= \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \: \gradpYlm - u^{\gradpY}_{l,m} \: \gradYlm ) ].
}
We can then write down the operator $\calR$ as follows:
\bseqn{
	\calR &=
		\begin{pmatrix}
			R & & & \\
			& R & & \\
			& & & \ddots
		\end{pmatrix} \in \R^{2(N+1)^2 \times 2(N+1)^2},
}
where the blocks $R$ are simply given by
\bseqn{
	R &:= 
		\begin{pmatrix}
			0 & 1 \\
			-1 & 0
		\end{pmatrix} \in \R^{2\times2}.
}
\end{proof}
